Derive final annotations based on NBAtlas reference

### 2025-08-19 analyses/cell-type-neuroblastoma-04/final-annotation.Rmd
add final annotation notebook, massively refreshed from prior versions

---
title: "Derive final annotations based on NBAtlas reference"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
params:
  scanvi_pp_threshold: 0.75    
  annotations_tsv: "results/SCPCP000004-annotations.tsv"
---

In this analysis module, we have annotated cell types in `SCPCP000004` samples using two methods: `SingleR` and `scANVI/scArches`, both using the `NBAtlas` dataset as a reference.
In this notebook, we derive the final annotations using information from the `SingleR` labels, the `scANVI/scArches` labels, and existing consensus cell type annotations.
All final annotations use labels present in the `NBAtlas` dataset.

@@

Next, we compare labels across all three methods: if at least _two out of three_ methods agree, we assign the agreeing label.
Note that this requires mapping consensus cell types to NBAtlas labels, which is only possible for these cell types: endothelial cells, fibroblasts, and immune cell types except for the `NBAtlas` labels `RBCs` and `pDC` which do not have corresponding consensus labels.

@@

* Cell type annotations can have directly corresponding labels. For example:
  * If the consensus cell type is `endothelial cell` and at least one of the `NBAtlas`-derived annotations is `Endothelial`, we assign the label `Endothelial`.
* Cell type annotations can differ but be in the same "family" of cell types. For example:
  * If the consensus cell type is `mature T cell` and at least one of the `NBAtlas`-derived annotations is `CD4+ T cell`, we assign the broader label `T cell`.
  * If one `NBAtlas`-derived annotation is `Patrolling monocyte` and the other is `Classical monocyte`, we assign the broader label `Monocyte`.

Finally, to the extent possible, we assign CL ontology IDs to the final annotations before export.

@@

theme_set(
  theme_bw()
)

@@

The function `get_unified_label()` defined in the chunk below determines a final label based on correspondence between at least two out of three labels, out of `SingleR`, `scANVI/scArches`, and the consensus label.

```{r}
#' Get unified cell type label
#'
#' @param annotation_df Data frame to annotate with columns `consensus`, `singler`, `scanvi`, and `cell_type_assignment`
#' @param consensus_options Vector of consensus cell types that match the final label, used for `consensus`
#' @param nbatlas_options Vector of NBAtlas cell types that match the final label, used for `singler` and `scanvi`
#' @param final_label Final label to apply if at least two methods have labels present in their respective vectors
#'
#' @returns Updated annotation_df with `cell_type_assignment` filled in with the `final_label` value where appropriate
get_unified_label <- function(
    annotation_df, 
    consensus_options, 
    nbatlas_options, 
    final_label) {

  annotation_df |>
    dplyr::mutate(
      # count how many annotations are in agreement
      n = consensus %in% consensus_options +
        singler %in% nbatlas_options + 
        scanvi %in% nbatlas_options, 
      # if >=2 agree, use this annotation
      cell_type_assignment = ifelse(
        is.na(cell_type_assignment) & n >= 2, 
        final_label, 
        cell_type_assignment
    )) |>
    dplyr::select(-n)
}
```

@@

```{r}
# Source additional utilities functions:
# - subset_nbatlas_markers()
# - harmonize_celltypes()
# - faceted_umap()
# - generate_dotplot()
source(file.path(module_dir, "scripts", "utils", "celltype-utils.R"))
```

@@

# marker genes from NBAtlas
nbatlas_markers_file <- file.path(
  ref_dir, 
  "nbatlas-marker-genes.tsv"
)
```

@@

Read SingleR results:

```{r}
singler_df <- singler_files |>
  purrr::map(readr::read_tsv) |>
  purrr::list_rbind(names_to = "library_id") |>
  dplyr::mutate(
    # add cell id column for unique rows & joining
    cell_id = glue::glue("{library_id}-{barcodes}"),
    # recode NA -> "Unknown"
    singler = ifelse(is.na(pruned.labels), "Unknown", pruned.labels)
  ) |>
  dplyr::select(cell_id, singler)
```

Read scANVI results:

```{r}
scanvi_df <- readr::read_tsv(
  file.path(scanvi_dir, "scanvi_predictions.tsv")
) |>
  dplyr::select(
    cell_id, 
    scanvi = scanvi_prediction, 
    starts_with("pp_")
  ) |>
  tidyr::pivot_longer(
    starts_with("pp_"), 
    names_to = "posterior_celltype", 
    values_to = "posterior"
  ) |>
  dplyr::mutate(posterior_celltype = stringr::str_remove(posterior_celltype, "^pp_")) |>
  dplyr::filter(scanvi == posterior_celltype) |>
  # recode to Unknown if below the threshold
  dplyr::mutate(scanvi = ifelse(posterior < params$scanvi_pp_threshold, "Unknown", scanvi)) |>
  dplyr::select(cell_id, scanvi)
```

@@

nbatlas_markers_df <- readr::read_tsv(nbatlas_markers_file)
```

@@

In this section, we'll derive the final annotations.

```{r}
ontology_slim_df <- ontology_df |>
  dplyr::select(-CL_annotation)

prep_for_annotation <- function(
    df, 
    annot_type, 
    ontology_df, 
    label_map_df) {

 df |>
    dplyr::select(all_of(c("cell_id", annot_type))) |>
    dplyr::rename(label = annot_type) |>
    # Join the family labels
    dplyr::left_join(label_map_df, by = c("label" = "NBAtlas_label_immune_zoom")) |>
    dplyr::rename(family_label = NBAtlas_label) |>
    # Ensure "Unknown" has a family label instead of NA for future-proofing
    dplyr::mutate(family_label = ifelse(is.na(family_label), "Unknown", family_label)) |>
    # Join in the ontology labels to match the **celltype** ontologies (label column)
    dplyr::left_join(ontology_slim_df, by = c("label" = "NBAtlas_label")) |>
    dplyr::rename(label_ontology = CL_ontology_id) |>
    # Join in the ontology labels to match the **family** ontologies (family_label column)
    dplyr::left_join(ontology_slim_df, by = c("family_label" = "NBAtlas_label")) |>
    dplyr::rename(family_ontology = CL_ontology_id) |>
    # We don't need the labels now that we have ontologies
    dplyr::select(-label, -family_label) |>
    # rename columns to start with `annot_type`
    dplyr::rename_with(\(x) {paste(annot_type, x, sep = "_")}, -cell_id)
}

singler_annotation_df <- prep_for_annotation(celltype_df, "singler", ontology_slim_df, label_map_df)
scanvi_annotation_df <- prep_for_annotation(celltype_df, "scanvi", ontology_slim_df, label_map_df)

annotation_df <- celltype_df |>
  dplyr::select(cell_id, consensus_ontology) |>
  dplyr::left_join(singler_annotation_df, by = "cell_id") |>
  dplyr::left_join(scanvi_annotation_df, by = "cell_id") |>
  dplyr::select(cell_id, consensus_ontology, starts_with("singler"), starts_with("scanvi"))
```

```{r}
annotation_df <- annotation_df |> 
  dplyr::mutate(
    CL_ontology_id = dplyr::case_when(
      # Use directly matching label
      scanvi_label_ontology == singler_label_ontology ~ scanvi_label_ontology, 
      # Check for family agreement between SingleR/scANVI
      scanvi_family_ontology == singler_family_ontology ~ scanvi_family_ontology,
      # Now use agreement with consensus to assign a label: first by label, and then by family
      singler_label_ontology == consensus_ontology ~ singler_label_ontology,
      scanvi_label_ontology == consensus_ontology ~ scanvi_label_ontology,
      singler_family_ontology == consensus_ontology ~ singler_family_ontology,
      scanvi_family_ontology == consensus_ontology ~ scanvi_family_ontology,
      # Everything else is NA
      .default = NA_character_
    )
  ) |>
  # keep only the final ontologies
  dplyr::select(cell_id, CL_ontology_id)

missing_cell_id <- annotation_df |>dplyr::filter(is.na(CL_ontology_id)) |>
  dplyr::pull(cell_id)

sjs_version <- readr::read_tsv(params$annotations_tsv)

stephyes_allyno <- sjs_version |>
  dplyr::filter(cell_id %in% missing_cell_id) |>
  dplyr::filter(!is.na(CL_ontology_id)) |>
  dplyr::pull(cell_id)

celltype_df |>
  dplyr::select(cell_id, singler, scanvi, consensus) |>
  dplyr::filter(cell_id %in% stephyes_allyno)
```

Next, we'll get unified labels for any cells where at least two of three methods agree.

```{r}
annotation_df <- annotation_df |>
  get_unified_label("fibroblast", "Fibroblast", "Fibroblast") |>
  get_unified_label("macrophage", "Macrophage", "Macrophage") |>
  get_unified_label("neutrophil", "Neutrophil", "Neutrophil") |>
  get_unified_label("plasma cell", "Plasma", "Plasma") |>
  get_unified_label(consensus_endo_celltypes, "Endothelial", "Endothelial") |>
  get_unified_label("natural killer cell", nbatlas_nk_celltypes, "NK cell") |>
  get_unified_label("monocyte", nbatlas_monocyte_celltypes, "Monocyte") |>
  get_unified_label("dendritic cell", nbatlas_cdc_celltypes, "cDC") |>
  get_unified_label(consensus_t_celltypes, nbatlas_t_celltypes, "T cell") |>
  # This must be last since it is the broadest category
  get_unified_label(consensus_myeloid_celltypes, nbatlas_myeloid_celltypes, "Myeloid")
```

@@

```{r}
sum(!is.na(annotation_df$CL_ontology_id)) / nrow(annotation_df)
```

@@

```{r, fig.width = 8, fig.height = 4}
frac_labeled_df <- annotation_df |>
  tidyr::separate(cell_id, into = c("library_id", "barcode"), remove = FALSE) |>
  dplyr::group_by(library_id) |>
  dplyr::summarize(
    frac_labeled = sum(!is.na(cell_type_assignment))/dplyr::n(), 
    library_size = dplyr::n()
  ) |>
  # get PDX logical for coloring
  dplyr::inner_join(sample_metadata, by = "library_id")

@@

```{r}
# Finalize the annotations
submission_df <- annotation_df |>
  # add ontology ids
  dplyr::left_join(
    nbatlas_ontology_ids, 
    by = c("cell_type_assignment" = "NBAtlas_label")
  ) |>
  # recode labels
  dplyr::mutate(
    cell_type_assignment = dplyr::case_when(
      is.na(cell_type_assignment) ~ "Unknown", 
      cell_type_assignment == "NE" ~ "Neuroendocrine/tumor",
      .default = cell_type_assignment
   ),
    tumor_cell_classification = dplyr::case_when(
      cell_type_assignment == "Unknown" ~ "Unknown", 
      cell_type_assignment == "Neuroendocrine/tumor" ~ "tumor", 
      .default = "normal"
    )
  ) |>
  tidyr::separate(cell_id, into = c("library_id", "cell_barcode"), remove = FALSE) |>
  dplyr::select(
    cell_id, library_id, cell_barcode, cell_type_assignment, tumor_cell_classification, CL_ontology_id
  )
# update palette label as well
names(celltype_pal)[names(celltype_pal) == "Neuroendocrine"] <- "Neuroendocrine/tumor"
```

```{r}
# prepare data frame for plotting
umap_df <- celltype_df |>
  dplyr::select(cell_id, sample_id, library_id, is_xenograft, UMAP1, UMAP2) |>
  dplyr::left_join(submission_df, by = "cell_id") |>
  # group "same family" cell types in the factor for cleaner legend
  # note this will give warnings if any of those cells aren't in the final annotation, but that's fine
  dplyr::mutate(
    cell_type_assignment = factor(cell_type_assignment), 
    cell_type_assignment = forcats::fct_relevel(
      cell_type_assignment, 
      "T cell", nbatlas_t_celltypes, 
      "NK cell", nbatlas_nk_celltypes,
      "cDC", nbatlas_cdc_celltypes, 
      "Monocyte", nbatlas_monocyte_celltypes
    )
  )
```

@@

Note that in the UMAP below, cell types in the same "family" each share a color:

* T cells: `T cell`, `Treg`, `CD4+ T cell`, `CD8+ T cell`
* Natural killer cells: `NK cell`, `Circulating NK cell`, `Resident NK cell`, `TOX2+/KIT+ NK cell`
* Conventional dendritic cells: `cDC`, `cDC1`, `cDC2/DC3`, `Migratory cDC`
* Monocytes: `Monocyte`, `Classical monocyte`, `Patrolling monocyte`

```{r fig.width = 10, fig.height = 10}
ggplot(umap_df) +
  aes(x = UMAP1, y = UMAP2, color = cell_type_assignment) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_color_manual(values = celltype_pal) +
guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
```

@@

This UMAP is the same as the previous one, except each facet highlights one specific cell type.

```{r fig.width = 16, fig.height = 16}
umap_theme <- list(
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(), 
    legend.position = "bottom", 
    aspect.ratio = 1
  )
)

umap_df |>
  faceted_umap(
    annotation_column = cell_type_assignment,
    celltype_colors = celltype_pal
  ) + 
  umap_theme
```

@@

```{r fig.width = 8, fig.height = 8}
ggplot(umap_df) +
  aes(x = UMAP1, y = UMAP2, color = tumor_cell_classification) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_color_brewer(palette = "Set2") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  umap_theme
```

@@

Marker genes that are present in three or more cell types are not included.
In addition, the upper limit for the color scale has been censored to a maximum of 6.

```{r}
# Temporary to develop code for next section
annotation_df <- readr::read_tsv(params$annotations_tsv) |>
  dplyr::filter(cell_type_assignment != "Unknown") |>
  # recode cell types to match what NBAtlas uses for their marker genes
  dplyr::mutate(
    dotplot_celltype = dplyr::case_when(
      stringr::str_detect(cell_type_assignment, "monocyte") ~ "Monocyte", 
      stringr::str_detect(cell_type_assignment, "cDC") ~ "cDC", 
      stringr::str_detect(cell_type_assignment, "Neuroendocrine") ~ "Neuroendocrine", 
      .default = cell_type_assignment
    )
  )

@@

```{r}
# get total number of cells per final annotation group and set up y_label
total_cells_df <- annotation_df |>
  dplyr::count(dotplot_celltype, name = "total_cells") |>
  # remove infrequent cell types
  dplyr::filter(total_cells > 10) |>
  dplyr::mutate(y_label = glue::glue("{dotplot_celltype} ({total_cells})")) |>
  # order by count, but put Unknown at the end
  dplyr::arrange(desc(total_cells)) |> 
  # amazing new dplyr trick; only this row is TRUE, so it gets moved to the last row
  dplyr::arrange(dotplot_celltype == "Unknown") |>
  # the dotplot function expects this column name
  dplyr::rename(label_recoded = dotplot_celltype)

@@

```{r, fig.width = 32, fig.height = 16}
generate_dotplot(
  merged_sce = merged_sce,
  markers_df = top_nbatlas_markers_df,
  # rename this after we've done all the wrangling
  celltype_df = annotation_df |> dplyr::rename(label_recoded = dotplot_celltype),
  total_cells_df = total_cells_df,
  expressed_genes = expressed_genes,
  bar_order = nbatlas_bar_order, 
  celltype_palette = celltype_pal, 
  min_cells = 0
) +
  theme(legend.position = "bottom")
```

@@

```{r}
readr::write_tsv(
  submission_df,
  params$annotations_tsv
)
```

>>>>

---
title: "Derive final annotations based on NBAtlas reference"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
params:
  scanvi_pp_threshold: 0.75    
  annotations_tsv: "../results/SCPCP000004-annotations.tsv"
---

In this analysis module, we have annotated cell types in `SCPCP000004` samples using two methods: `SingleR` and `scANVI/scArches`, both using the NBAtlas dataset as a reference.
In this notebook, we derive the final annotations using information from the `SingleR` labels, the `scANVI/scArches` labels, and existing consensus cell type annotations.
All final annotations use labels present in the NBAtlas dataset.

@@

Next, we compare labels across all three methods: if at least _two out of three_ methods agree, we assign the agreeing label.
Note that this requires mapping consensus cell types to NBAtlas labels, which is only possible for these cell types: endothelial cells, fibroblasts, and immune cell types except for the NBAtlas labels `RBCs` and `pDC` which do not have corresponding consensus labels.

@@

* Cell type annotations can have directly corresponding labels. For example:
  * If the consensus cell type is `endothelial cell` and at least one of the NBAtlas-derived annotations is `Endothelial`, we assign the label `Endothelial`.
* Cell type annotations can differ but be in the same "family" of cell types. For example:
  * If the consensus cell type is `mature T cell` and at least one of the NBAtlas-derived annotations is `CD4+ T cell`, we assign the broader label `T cell`.
  * If one NBAtlas-derived annotation is `Patrolling monocyte` and the other is `Classical monocyte`, we assign the broader label `Monocyte`.

We use CL ontology ids to determine agreement throughout. 
For more information about how these ontology ids were determined for NBAtlas labels, see `../references/README.md`.

For the final exported TSV file, we further designate any cells labeled as `Neuroendocrine` as `tumor`, following how this label is interpreted in NBAtlas.

@@

theme_set(
  theme_bw()
)
umap_theme <- list(
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(), 
    legend.position = "bottom", 
    aspect.ratio = 1
  )
)

@@

@@

@@

# marker genes from NBAtlas
nbatlas_markers_file <- file.path(
  ref_dir, 
  "nbatlas-marker-genes.tsv"
)

# validation groups URL, used to obtain consensus "family" ontologies (aka broad validation group ontologies)
validation_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-validation-groups.tsv"
```

```{r}
# Source utilities functions:
# - subset_nbatlas_markers()
# - faceted_umap()
# - generate_dotplot()
source(file.path(module_dir, "scripts", "utils", "celltype-utils.R"))
```

```{r}

#' Prepare data frame of scANVI or SingleR labels for annotation
#'
#' @param df Data frame to prepare
#' @param annot_type "singler" or "scanvi"
#' @param ontology_df Data frame of ontology ids
#' @param label_map_df Data frame mapping labels to family labels
#'
#' @returns Wide data frame with ontologies and family labels for the given annotation type
prep_for_annotation <- function(
    df, 
    annot_type, 
    ontology_df, 
    label_map_df) {

  df |>
    dplyr::select(all_of(c("cell_id", annot_type))) |>
    dplyr::rename(label = annot_type) |>
    ####### Join in the family labels
    dplyr::left_join(label_map_df, by = c("label" = "NBAtlas_label")) |>
    dplyr::rename(family = NBAtlas_family) |>
    ######### Obtain LABEL ontologies
    dplyr::left_join(ontology_slim_df, by = c("label" = "NBAtlas_label")) |>
    dplyr::rename(label_ontology = CL_ontology_id) |>
    ######## Obtain FAMILY ontologies
    dplyr::left_join(ontology_slim_df, by = c("family" = "NBAtlas_label")) |>
    dplyr::rename(family_ontology = CL_ontology_id) |>
    # rename columns to start with `annot_type`
    dplyr::rename_with(\(x) {paste(annot_type, x, sep = "_")}, -cell_id) 
}
```

@@

Read `SingleR` results:

```{r}
singler_df <- singler_files |>
  purrr::map(readr::read_tsv) |>
  purrr::list_rbind(names_to = "library_id") |>
  dplyr::mutate(
    # add cell id column for unique rows & joining
    cell_id = glue::glue("{library_id}-{barcodes}"),
    # recode NA -> "Unknown" and NE -> "Neuroendocrine"
    singler = dplyr::case_when(
      is.na(pruned.labels) ~ "Unknown", 
      pruned.labels == "NE" ~ "Neuroendocrine",
      .default = pruned.labels)
  ) |>
  dplyr::select(cell_id, singler)
```

Read `scANVI` results:

```{r}
scanvi_df <- readr::read_tsv(
  file.path(scanvi_dir, "scanvi_predictions.tsv")
) |>
  dplyr::select(
    cell_id, 
    scanvi = scanvi_prediction, 
    starts_with("pp_")
  ) |>
  tidyr::pivot_longer(
    starts_with("pp_"), 
    names_to = "posterior_celltype", 
    values_to = "posterior"
  ) |>
  dplyr::mutate(posterior_celltype = stringr::str_remove(posterior_celltype, "^pp_")) |>
  dplyr::filter(scanvi == posterior_celltype) |>
  # recode to Unknown if below the threshold, and NE -> Neuroendocrine
  dplyr::mutate(scanvi = dplyr::case_when(
    posterior < params$scanvi_pp_threshold ~ "Unknown", 
    scanvi == "NE" ~ "Neuroendocrine", 
    .default = scanvi)) |>
  dplyr::select(cell_id, scanvi)
```

@@

nbatlas_markers_df <- readr::read_tsv(nbatlas_markers_file)

validation_df <- readr::read_tsv(validation_url) |>
  dplyr::rename(
    consensus = consensus_annotation,
    consensus_family_label = validation_group_annotation,
    consensus_family_ontology = validation_group_ontology 
  )
```

@@

Firt we prepare the data frame for annotation.

```{r, warning = FALSE}
ontology_slim_df <- ontology_df |>
  dplyr::select(-CL_annotation)

singler_annotation_df <- prep_for_annotation(
  celltype_df, 
  "singler", 
  ontology_slim_df, 
  label_map_df
)
scanvi_annotation_df <- prep_for_annotation(
  celltype_df, 
  "scanvi", 
  ontology_slim_df, 
  label_map_df
)

annotation_df <- celltype_df |>
  dplyr::select(cell_id, consensus, consensus_ontology) |>
  dplyr::left_join(singler_annotation_df, by = "cell_id") |>
  dplyr::left_join(scanvi_annotation_df, by = "cell_id") |>
  dplyr::left_join(validation_df, by = c("consensus", "consensus_ontology")) |>
  dplyr::select(cell_id, starts_with("consensus"), starts_with("singler"), starts_with("scanvi"))
```

In the code below, we compare ontology ids to derive final annotations.
However, based on those comparisons, we assign a final _label_, not final ontology id, and then add the final ontology ids back into those annotations.
This is because we have several `NA` ontology ids, and we would not be able to unambiguously map their labels in if we assigned ontologies.

```{r}
final_annotation_df <- annotation_df |> 
  dplyr::mutate(
    # For all comparisons, make sure we aren't comparing NA to NA and getting TRUE
    final_label = dplyr::case_when(
      ########### Check for exact match between SingleR/scANVI
      !is.na(scanvi_label_ontology) & singler_label_ontology == scanvi_label_ontology ~ scanvi_label, 
      ########### Check for family match between SingleR/scANVI
      !is.na(scanvi_family_ontology) & singler_family_ontology == scanvi_family_ontology ~ scanvi_family,
      ########## Now use agreement with consensus to assign a label: first by label, and then by family
      !is.na(consensus_ontology) & singler_label_ontology  == consensus_ontology ~ singler_label,
      !is.na(consensus_ontology) & scanvi_label_ontology   == consensus_ontology ~ scanvi_label,
      !is.na(consensus_family_ontology) & singler_family_ontology == consensus_family_ontology ~ singler_family,
      !is.na(consensus_family_ontology) & scanvi_family_ontology  == consensus_family_ontology ~ scanvi_family,
      # Everything else is Unknown
      .default = "Unknown"
    )
  ) |>
  # Now, we can bring map the final ontologies into the data frame
  dplyr::inner_join(
    ontology_df |> dplyr::rename(final_label = NBAtlas_label, final_id = CL_ontology_id), 
    by = "final_label"
  ) |>
  # add tumor column for visualization
  dplyr::mutate(cell_class = dplyr::case_when(
    final_label == "Neuroendocrine" ~ "tumor", 
    final_label == "Unknown" ~ "unknown", 
    .default = "normal"
  ))
```

@@

```{r}
sum(!is.na(final_annotation_df$final_id)) / nrow(final_annotation_df)
```

@@

```{r, fig.width = 8, fig.height = 4}
frac_labeled_df <- final_annotation_df |>
  tidyr::separate(cell_id, into = c("library_id", "barcode"), remove = FALSE) |>
  dplyr::group_by(library_id) |>
  dplyr::summarize(
    frac_labeled = sum(!is.na(final_id))/dplyr::n(), 
    library_size = dplyr::n()
  ) |>
  # get PDX logical for coloring
  dplyr::inner_join(sample_metadata, by = "library_id")

@@

```{r}
label_order <- label_map_df |>
  dplyr::add_count(NBAtlas_family) |>
  dplyr::arrange(desc(n), NBAtlas_family) |>
  dplyr::filter(NBAtlas_label %in% final_annotation_df$final_label) |>
  dplyr::pull(NBAtlas_label)
label_order <- c(label_order, "Unknown")

# prepare data frame for plotting
umap_df <- celltype_df |>
  dplyr::select(cell_id, sample_id, library_id, is_xenograft, UMAP1, UMAP2) |>
  dplyr::left_join(final_annotation_df, by = "cell_id") |>
  dplyr::mutate(final_label = factor(final_label, levels = label_order))
```

@@

Note that in the UMAP below, cell types in these same "family" each share a color:

* T cells
* Natural killer cells
* Conventional dendritic cells
* Monocytes

```{r fig.width = 10, fig.height = 10}
ggplot(umap_df) +
  aes(x = UMAP1, y = UMAP2, color = final_label) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_color_manual(values = celltype_pal) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) + 
  umap_theme +
  theme(legend.position = "bottom")
```

@@

This UMAP is the same as the previous one, except each facet highlights one specific cell type.
In this plot, only cell types with at least 10 cells are shown.

```{r fig.width = 16, fig.height = 16}
umap_df |>
  dplyr::add_count(final_label) |>
  dplyr::filter(n > 10) |>
  faceted_umap(
    annotation_column = final_label,
    celltype_colors = celltype_pal, 
    facet_wrap_cols = 6
  ) + 
    umap_theme +
    theme(legend.position = "none")
```

@@

```{r fig.width = 8, fig.height = 8}
ggplot(umap_df) +
  aes(x = UMAP1, y = UMAP2, color = cell_class) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_color_brewer(palette = "Set2") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  umap_theme
```

@@

Marker genes that are present in three or more cell types are not included.
In addition, the upper limit for the color scale has been censored to a maximum of 6 to ensure visibility of expression levels across all cell types.

```{r}
input_dotplot_df <- final_annotation_df |>
  dplyr::filter(final_label != "Unknown") |>
  # recode cell types to match what is present in NBAtlas marker genes
  dplyr::mutate(
    label_recoded = dplyr::case_when(
      stringr::str_detect(final_label, "monocyte") ~ "Monocyte", 
      stringr::str_detect(final_label, "cDC") ~ "cDC",
      .default = final_label
  )) |>
  dplyr::select(cell_id, label_recoded)

@@

```{r}
# get total number of cells per final annotation group and set up y_label
total_cells_df <- input_dotplot_df |>
  dplyr::count(label_recoded, name = "total_cells") |>
  dplyr::mutate(y_label = glue::glue("{label_recoded} ({total_cells})")) |>
  dplyr::inner_join(label_map_df, by = c("label_recoded" = "NBAtlas_label")) |>
  dplyr::group_by(NBAtlas_family) |>
  dplyr::mutate(family_total_cells = sum(total_cells)) |>
  # order by family count, but put Unknown at the end
  dplyr::arrange(desc(family_total_cells)) |>
  # amazing new dplyr trick; only this row is TRUE, so it gets moved to the last row
  dplyr::arrange(label_recoded == "Unknown")

@@

```{r, fig.width = 32, fig.height = 16}
generate_dotplot(
  merged_sce = merged_sce,
  markers_df = top_nbatlas_markers_df,
  celltype_df = input_dotplot_df,
  total_cells_df = total_cells_df,
  expressed_genes = expressed_genes,
  bar_order = nbatlas_bar_order, 
  celltype_palette = celltype_pal, 
  min_cells = 10 # only show cell types with at least 10 cells
) +
  theme(legend.position = "bottom")
```

@@

Export the final table with submission-ready column names.

```{r}
final_annotation_df |>
  dplyr::select(
    cell_id, 
    cell_type_assignment = final_label, 
    tumor_cell_classification = cell_class,
    CL_ontology_id = final_id, 
    CL_annotation
  ) |>
  readr::write_tsv(
    params$annotations_tsv
  )
```
